---
title: "Explaining roe deer movement in the Italian alps"
authors: Anne-Juul Welsink and Marrit Leenstra
date: 31 January 2019
output: html_notebook
---

TODO:
Project description

```{r}
# check if libraries are installed
if (!require("dplyr")) install.packages("dplyr")
if (!require("sp")) install.packages("sp")
if (!require("sf")) install.packages("sf")
if (!require("DBI")) install.packages("DBI")
if (!require("RSQLite")) install.packages("RSQLite")
if (!require("raster")) install.packages("raster")
if (!require("adehabitatLT")) install.packages("adehabitatLT")
if (!require("lubridate")) install.packages("lubridate")
if (!require("ggplot2")) install.packages("ggplot2")
```

```{r}
# load libraries
library(dplyr)
library(sp)
library(sf)
library(DBI)
library(raster)
library(adehabitatLT)
library(lubridate)
library(ggplot2)
```

```{r}
# import functions
source("R/retrieveData.R")
source("R/saveDataInDB.R")
```

```{r}
# project variables
movementDataURL = "extras.springer.com/2014/978-3-319-03742-4/trackingDB_datasets.zip"
environmentalURL = "https://drive.google.com/uc?export=download&id=18LX2hmiYEgcdO1SfzhfzItHx3ii1WKKH"
```

```{r}
# download movement data
folder = "data"
retrieveData(movementDataURL, folder)

# download environmental data
retrieveData(environmentalURL, folder)
```

```{r}
# Load environmental data to global environment
fileList <- list.files(path = "data/GoogleEarthData", pattern = glob2rx('*.tif'), full.names = TRUE)
DEM <- raster(fileList[1])
landUse <- raster(fileList[2])
slope <- raster(fileList[3])
```

We use an example dataset of roe deer movement from a book about spatial database management (Urbano and Cagnacci, 2014). The data are thus designed for use in a database and set up as interlinking datasets. Direct use in R is possible, but undesirable due to a lack of structure. Therefore, we set up an SQLite database in memory to process the data.

```{r}
# create database connection
movementDB <- dbConnect(RSQLite::SQLite(), "")

# get files
gpsData <- list.files(path = "data/tracking_db/data/sensors_data/", pattern = glob2rx('GSM0*.csv'), full.names = TRUE)
sensorsAnimals <- list.files(path = "data/tracking_db/data/sensors_animals/", pattern = glob2rx('*.csv'), full.names = TRUE)
sensors <- list.files(path = "data/tracking_db/data/sensors/", pattern = glob2rx('*.csv'), full.names = TRUE)
animals <- list.files(path = "data/tracking_db/data/animals/", pattern = glob2rx('*.csv'), full.names = TRUE)

# create dataframe

# specify column names for database tables
gpsDataColumnNames <- c("gpsSensorsCode","lineNo", "utcDate", "utcTime", "lmtDate", "lmtTime", "ecefX", "ecefY","ecefZ", "latitude", "longitude", 
"height", "dop", "nav", "validated", "satsUsed", "ch01SatId", "ch01SatCnr", "ch02SatId", "ch02SatCnr", "ch03SatId", "ch03SatCnr", "ch04SatId", "ch04SatCnr", "ch05SatId", "ch05SatCnr", "ch06SatId", "ch06SatCnr", "ch07SatId", "ch07SatCnr", "ch08SatId", "ch08SatCnr", "ch09SatId", "ch09SatCnr", "ch10SatId", "ch10SatCnr", "ch11SatId", "ch11SatCnr", "ch12SatId", "ch12SatCnr", "mainVol", "buVol", "temp", "easting", "northing", "remarks")
sensorsAnimalsColumnNames <- c("animalsId","gpsSensorsId","startTime","EndTime","notes")
sensorsColumnNames <-c("gpsSensorsId", "gpsSensorsCode", "purchaseDate", "frequency", "vendor","model","sim")
animalsColumnNames <- c("animalsId", "animalsCode", "name","sex","ageClassCode", "speciesCode")

# save data in database
saveDataInDB(gpsData, movementDB, "gpsData", gpsDataColumnNames, TRUE)
saveDataInDB(sensorsAnimals, movementDB, "sensorsAnimals", sensorsAnimalsColumnNames, FALSE)
saveDataInDB(sensors, movementDB, "sensors", sensorsColumnNames, FALSE)
saveDataInDB(animals, movementDB, "animals", animalsColumnNames, FALSE)
```

```{r}
# use data
#dbListTables(movementDB)
dbListFields(movementDB, "gpsData")
dbListFields(movementDB, "sensorsAnimals")
dbListFields(movementDB, "sensors")
dbListFields(movementDB, "animals")
#dbGetQuery(movementDB, 'SELECT "GPSSensorsCode", "latitude", "longitude" FROM GPSDATA WHERE #"GPSSensorsCode" in ("GSM01508", "GSM01438")')
```

The database can be queried with SQL. Since we want to link the GPS data with specific animals we need to join these tables. However, these tables do not have any common fields that can be used for a straightforward join. Embedding of SQL statements allows us to get to the required combination of data. 

```{r}
# query data from database
query <- '
SELECT
animals.name,
animals.sex,
dataWithAnimalId.utcDate, dataWithAnimalId.utcTime, 
dataWithAnimalId.latitude, dataWithAnimalId.longitude,
dataWithAnimalId.height
FROM
  (SELECT 
  dataWithSensorId.gpsSensorsCode, 
  dataWithSensorId.utcDate, dataWithSensorId.utcTime, 
  dataWithSensorId.latitude, dataWithSensorId.longitude, 
  dataWithSensorId.height,
  dataWithSensorId.gpsSensorsId, 
  sensorsAnimals.animalsId 
  FROM 
    (SELECT 
    gpsData.gpsSensorsCode, 
    gpsData.utcDate, gpsData.utcTime, 
    gpsData.latitude, gpsData.longitude, 
    gpsData.height,
    sensors.gpsSensorsId 
    FROM 
    gpsData 
    LEFT JOIN 
    sensors ON gpsData.gpsSensorsCode = sensors.gpsSensorsCode) AS dataWithSensorId 
  LEFT JOIN 
  sensorsAnimals ON dataWithSensorId.gpsSensorsId = sensorsAnimals.gpsSensorsId) AS dataWithAnimalId 
LEFT JOIN
animals ON dataWithAnimalId.animalsId = animals.animalsId
;'

selectedMovementData <- dbGetQuery(movementDB, query)
```

At this point we have the data we need. However, all data is stored as strings. The correct classes are assigned, before we can create a geometry out of the latitude and longitude columns. We choose the WGS84 zone 32 (EPSG:32632) as coordinate system as it provides enough detail for our analyses. 

```{r}
# correct classes
selectedMovementData$sex <- factor(selectedMovementData$sex)
class(selectedMovementData$latitude) <- "numeric"
class(selectedMovementData$longitude) <- "numeric"
class(selectedMovementData$height) <- "numeric"
selectedMovementData$date <- dmy_hms(paste(selectedMovementData$utcDate, selectedMovementData$utcTime), tz = "UTC")
# remove redundant columns
selectedMovementData[3:4] <- list(NULL)

# remove rows with NA for location
selectedMovementDataNoNA <- selectedMovementData[!is.na(selectedMovementData$latitude) | !is.na(selectedMovementData$longitude),]

# create spatial data frame projected in WGS84 zone 32
coords <- selectedMovementDataNoNA[,c(4:3)]
proj4WGS84 <- "+proj=longlat +datum=WGS84 +no_defs"
proj4 <- "+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs"
WGS84movementSDF <- SpatialPointsDataFrame(coords = coords, data = selectedMovementDataNoNA, proj4string = CRS(proj4WGS84))
movementSDF <- spTransform(WGS84movementSDF, CRS(proj4))

# create trajectories
movementTrajectories <- as.ltraj(xy = coordinates(movementSDF), date = movementSDF$date, id = movementSDF$name)

# reproject rasters
DEM <- projectRaster(DEM, crs=proj4)
```

We are curious where our roe deer are going! 

```{r}
# check data
plot(DEM)
lines(movementTrajectories[[1]])
lines(movementTrajectories[[2]], col = 'red')
lines(movementTrajectories[[3]], col = 'blue')
lines(movementTrajectories[[4]], col = 'purple')
lines(movementTrajectories[[5]], col = 'orange')

par(mfrow=c(3, 2))
hist(movementTrajectories[[1]]$dist, xlim = c(0, 2000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[1]))
hist(movementTrajectories[[2]]$dist, xlim = c(0, 2000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[2]))
hist(movementTrajectories[[3]]$dist, xlim = c(0, 2000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[3]))
hist(movementTrajectories[[4]]$dist, xlim = c(0, 2000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[4]))
hist(movementTrajectories[[5]]$dist, xlim = c(0, 2000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[5]))

# quantiles give a good indication of the distribution (around 4h between relocations)
quantile(diff(movementTrajectories[[1]][,'date']))
quantile(diff(movementTrajectories[[2]][,'date']))
quantile(diff(movementTrajectories[[3]][,'date']))
quantile(diff(movementTrajectories[[4]][,'date']))
quantile(diff(movementTrajectories[[5]][,'date']))

par(mfrow=c(1, 1))
hist(as.numeric(diff(movementTrajectories[[1]][,'date'])), xlim = c(0,5000), ylim = c(0,1000), breaks = seq(0, 10000000, by = 100), xlab = "time", main = paste("Time between relocations by" , id(movementTrajectories[1])))
hist(as.numeric(diff(movementTrajectories[[2]][,'date'])), xlim = c(0,5000), ylim = c(0,1000), breaks = seq(0, 10000000, by = 100), main = paste("Time between relocations by" , id(movementTrajectories[2])))

dataframeTrajectories <- ld(movementTrajectories)
groups <- dataframeTrajectories %>% group_by(id) %>% mutate(timeDifference = c(NA, diff(date)))
trajectories <- dl(groups)

# boxplots to find best variable to remove outliers
par(mfrow=c(3,2))
ggplot(groups, aes(x=id, y=dist, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groups, aes(x=id, y=timeDifference, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groups, aes(x=id, y=rel.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groups, aes(x=id, y=abs.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groups, aes(x=id, y=y, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groups, aes(x=id, y=x, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
```

The data clearly contain some outliers. These are removed in order to perform reliable analyses. ################# Uitleg van cut-off point? Outliers are identified on the basis on the distance between relocations. A moved distance of .... or higher is considered unrealistic. #######################

```{r}
# clean data
# remove outliers based on x-coordinates and distance between relocations
groupsCleaned <- groups %>% filter(x < 700000) %>% filter(x > 650000) %>% filter(dist < 000)

# cast back to ltraj 
movementTrajectoriesCleaned <- dl(groupsCleaned)

# check data again
plot(DEM)
lines(movementTrajectoriesCleaned[[1]])
lines(movementTrajectoriesCleaned[[2]], col = 'red')
lines(movementTrajectoriesCleaned[[3]], col = 'blue')
lines(movementTrajectoriesCleaned[[4]], col = 'purple')
lines(movementTrajectoriesCleaned[[5]], col = 'orange')

# boxplots
ggplot(groupsCleaned, aes(x=id, y=dist, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groupsCleaned, aes(x=id, y=timeDifference, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groupsCleaned, aes(x=id, y=y, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")
ggplot(groupsCleaned, aes(x=id, y=x, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none")

```



```{r}


```

```{r}
# disconnect database
dbDisconnect(movementDB)
```

## References
Urbano, F., & Cagnacci, F. (2014). Spatial database for GPS wildlife tracking data. Springer International Publishing.