---
title: "Explaining roe deer movement in the Italian alps"
authors: Anne-Juul Welsink and Marrit Leenstra
date: 31 January 2019
output: html_notebook
---

TODO:
Project description

```{r}
# check if libraries are installed
if (!require("dplyr")) install.packages("dplyr")
if (!require("sp")) install.packages("sp")
if (!require("sf")) install.packages("sf")
if (!require("DBI")) install.packages("DBI")
if (!require("RSQLite")) install.packages("RSQLite")
if (!require("raster")) install.packages("raster")
if (!require("adehabitatLT")) install.packages("adehabitatLT")
if (!require("lubridate")) install.packages("lubridate")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("leaflet")) install.packages("leaflet")
if (!require("rlist")) install.packages("rlist")
if (!require("entropy")) install.packages("entropy")
```

```{r}
# load libraries
library(dplyr)
library(sp)
library(sf)
library(DBI)
library(raster)
library(adehabitatLT)
library(lubridate)
library(ggplot2)
library(gridExtra)
library(leaflet)
library(rlist)
library(entropy)
```

```{r}
# import functions
source("R/retrieveData.R")
source("R/saveDataInDB.R")
source("R/simulateMovement.R")
source("R/createSpatialLine.R")
source("R/extractRasterValues.R")
source("R/plotProfiles.R")
source("R/combineSimTrack.R")
source("R/sumStats.R")
source("R/boxplotProfiles.R")
source("R/addLegend.R")
```

```{r}
# project variables

# downloadlinks
movementDataURL = "extras.springer.com/2014/978-3-319-03742-4/trackingDB_datasets.zip"
environmentalURL = "https://drive.google.com/uc?export=download&id=1nb7q8TQGAkzTcGtSMS0Nig_DAhB2lClJ"

landUseLegendURL  = "https://www.eea.europa.eu/data-and-maps/data/corine-land-cover-2006-raster-1/corine-land-cover-classes-and/clc_legend.csv/at_download/file"

# projections
proj4WGS84 <- "+proj=longlat +datum=WGS84 +no_defs"
proj4zone32 <- "+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs"
proj4Mercator <- "+proj=merc +a=6378137 +b=6378137 +lat_ts=0.0 +lon_0=0.0 +x_0=0.0 +y_0=0 +k=1.0 +units=m +nadgrids=@null +wktext  +no_defs"
```

```{r}
# download movement data and unzip
folder = "data"
retrieveData(movementDataURL, folder, unzip = TRUE)

# download environmental data and unzip
retrieveData(environmentalURL, folder, unzip = TRUE)

# download legend for corine land cover (landUse)
retrieveData(landUseLegendURL, folder, filename = "landUseLegend.csv")
```

```{r}
# Load environmental data to global environment
DEM <- raster(list.files(path = "data/GoogleEarthData", pattern = glob2rx('*DEM*.tif'), full.names = TRUE))
landsat <- brick(list.files(path = "data/GoogleEarthData", pattern = glob2rx('*Landsat*.tif'), full.names = TRUE))
landUse <- raster(list.files(path = "data/GoogleEarthData", pattern = glob2rx('*LandUse*.tif'), full.names = TRUE))
slope <- raster(list.files(path = "data/GoogleEarthData", pattern = glob2rx('*Slope*.tif'), full.names = TRUE))
landUseLegend <- read.csv(list.files(path = folder, pattern = glob2rx('*.csv'), full.names = TRUE))
```

We use an example dataset of roe deer movement from a book about spatial database management (Urbano and Cagnacci, 2014). The data are thus designed for use in a database and set up as interlinking datasets. Direct use in R is possible, but undesirable due to a lack of structure. Therefore, we set up an SQLite database in memory to process the data.

```{r}
# create database connection
movementDB <- dbConnect(RSQLite::SQLite(), "")

# get files
gpsData <- list.files(path = "data/tracking_db/data/sensors_data/", pattern = glob2rx('GSM0*.csv'), full.names = TRUE)
sensorsAnimals <- list.files(path = "data/tracking_db/data/sensors_animals/", pattern = glob2rx('*.csv'), full.names = TRUE)
sensors <- list.files(path = "data/tracking_db/data/sensors/", pattern = glob2rx('*.csv'), full.names = TRUE)
animals <- list.files(path = "data/tracking_db/data/animals/", pattern = glob2rx('*.csv'), full.names = TRUE)

# specify column names for database tables
gpsDataColumnNames <- c("gpsSensorsCode","lineNo", "utcDate", "utcTime", "lmtDate", "lmtTime", "ecefX", "ecefY","ecefZ", "latitude", "longitude", 
"height", "dop", "nav", "validated", "satsUsed", "ch01SatId", "ch01SatCnr", "ch02SatId", "ch02SatCnr", "ch03SatId", "ch03SatCnr", "ch04SatId", "ch04SatCnr", "ch05SatId", "ch05SatCnr", "ch06SatId", "ch06SatCnr", "ch07SatId", "ch07SatCnr", "ch08SatId", "ch08SatCnr", "ch09SatId", "ch09SatCnr", "ch10SatId", "ch10SatCnr", "ch11SatId", "ch11SatCnr", "ch12SatId", "ch12SatCnr", "mainVol", "buVol", "temp", "easting", "northing", "remarks")
sensorsAnimalsColumnNames <- c("animalsId","gpsSensorsId","startTime","EndTime","notes")
sensorsColumnNames <-c("gpsSensorsId", "gpsSensorsCode", "purchaseDate", "frequency", "vendor","model","sim")
animalsColumnNames <- c("animalsId", "animalsCode", "name","sex","ageClassCode", "speciesCode")

# save data in database
saveDataInDB(gpsData, movementDB, "gpsData", gpsDataColumnNames, TRUE)
saveDataInDB(sensorsAnimals, movementDB, "sensorsAnimals", sensorsAnimalsColumnNames, FALSE)
saveDataInDB(sensors, movementDB, "sensors", sensorsColumnNames, FALSE)
saveDataInDB(animals, movementDB, "animals", animalsColumnNames, FALSE)
```

The database can be queried with SQL. Since we want to link the GPS data with specific animals we need to join these tables. However, these tables do not have any common fields that can be used for a straightforward join. Embedding of SQL statements allows us to get to the required combination of data. 

```{r}
# query data from database
query <- '
SELECT
animals.name,
animals.sex,
dataWithAnimalId.utcDate, dataWithAnimalId.utcTime, 
dataWithAnimalId.latitude, dataWithAnimalId.longitude,
dataWithAnimalId.height
FROM
  (SELECT 
  dataWithSensorId.gpsSensorsCode, 
  dataWithSensorId.utcDate, dataWithSensorId.utcTime, 
  dataWithSensorId.latitude, dataWithSensorId.longitude, 
  dataWithSensorId.height,
  dataWithSensorId.gpsSensorsId, 
  sensorsAnimals.animalsId 
  FROM 
    (SELECT 
    gpsData.gpsSensorsCode, 
    gpsData.utcDate, gpsData.utcTime, 
    gpsData.latitude, gpsData.longitude, 
    gpsData.height,
    sensors.gpsSensorsId 
    FROM 
    gpsData 
    LEFT JOIN 
    sensors ON gpsData.gpsSensorsCode = sensors.gpsSensorsCode) AS dataWithSensorId 
  LEFT JOIN 
  sensorsAnimals ON dataWithSensorId.gpsSensorsId = sensorsAnimals.gpsSensorsId) AS dataWithAnimalId 
LEFT JOIN
animals ON dataWithAnimalId.animalsId = animals.animalsId
;'

selectedMovementData <- dbGetQuery(movementDB, query)
```

At this point we have the data we need. However, all data is stored as strings. The correct classes are assigned, before we can create a geometry out of the latitude and longitude columns. We choose the WGS84 zone 32 (EPSG:32632) as coordinate system as it provides enough detail for our analyses. 

```{r}
# correct classes
selectedMovementData$sex <- factor(selectedMovementData$sex)
class(selectedMovementData$latitude) <- "numeric"
class(selectedMovementData$longitude) <- "numeric"
class(selectedMovementData$height) <- "numeric"
selectedMovementData$date <- dmy_hms(paste(selectedMovementData$utcDate, selectedMovementData$utcTime), tz = "UTC")
# remove redundant columns
selectedMovementData[3:4] <- list(NULL)

# remove rows with NA for location
selectedMovementDataNoNA <- selectedMovementData[!is.na(selectedMovementData$latitude) | !is.na(selectedMovementData$longitude),]

# create spatial data frame projected in WGS84 zone 32
coords <- selectedMovementDataNoNA[,c(4:3)]
WGS84movementSDF <- SpatialPointsDataFrame(coords = coords, data = selectedMovementDataNoNA, proj4string = CRS(proj4WGS84))
movementSDF <- spTransform(WGS84movementSDF, CRS(proj4zone32))

# create trajectories
movementTrajectories <- as.ltraj(xy = coordinates(movementSDF), date = movementSDF$date, id = movementSDF$name)
```

We are curious where "our" roe deer are going! 
For visualization purposes we need to create spatial lines of the GPS recorded trajectories.

```{r}
# select distinct animals
animalNames <- distinct(groups, id)[[1]]

for (animal in animalNames){
  # get location data from nth animal
  locationdatainit <- groups[(groups$id == animal),]
  
  # create spatial line
  assign(paste0("trajectinit",animal), createSpatialLine(locationdata = locationdatainit, proj4dataframe = proj4zone32, proj4line = proj4WGS84, lineId = animal))
}

```

An interactive leaflet map allows us to explore the GPS recorded movement of our roe deer, Agostino, Alessandra, Daniela, Decimo, and Sandro.
```{r}
m <- leaflet() %>% 
  setView(lng=11.0367, lat=46.0331, zoom = 11) %>%
  # baselayers
  addTiles(group = 'Street map') %>%  
  addProviderTiles('Stamen.Terrain', group = 'Terrain') %>% 
  addProviderTiles('Esri.WorldImagery', group = 'Satellite') %>%
  # overlay groups
  addPolylines(data = trajectinitAgostino, weight = 1, opacity = 0.8, color = "red", group = 'Agostino') %>% 
  addPolylines(data = trajectinitAlessandra, weight = 1, opacity = 0.8, color = "blue", group = 'Alessandra') %>% 
  addPolylines(data = trajectinitDaniela, weight = 1, opacity = 0.8, color = "orange", group = 'Daniela') %>% 
  addPolylines(data = trajectinitDecimo, weight = 1, opacity = 0.8, color = "yellow", group = 'Decimo') %>% 
  addPolylines(data = trajectinitSandro, weight = 1, opacity = 0.8, color = "purple", group = 'Sandro') %>% 
  addLayersControl(
    baseGroups = c('Street map', 'Terrain', 'Satellite'),
    overlayGroups = c('Agostino', 'Alessandra', 'Daniela', 'Decimo', 'Sandro'),
    options = layersControlOptions(collapsed = FALSE)
  )
m 

```

Oops! The data clearly contain some outliers. These need to be removed in order to perform reliable analyses. But how do we define which points are outliers? We look at boxplots of different variables in order to determine a suitable method to identify outliers. 

```{r}
# cast back to dataframe
dataframeTrajectories <- ld(movementTrajectories)
groups <- dataframeTrajectories %>% mutate(height = movementSDF$height) %>% group_by(id)
trajectories <- dl(groups)

# plot
p1 <- ggplot(groups, aes(x=id, y=x, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("x-coordinates of locations")
p2 <- ggplot(groups, aes(x=id, y=y, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("y-coordinates of locations")
p3 <- ggplot(groups, aes(x=id, y=dist, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Distance between relocations")
p4 <- ggplot(groups, aes(x=id, y=abs.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Absolute angle between relocations")
p5 <- ggplot(groups, aes(x=id, y=rel.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Relative angle between relocations")
p6 <- ggplot(groups, aes(x=id, y=dt, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Time between relocations")

grid.arrange(p1,p2,p3,p4,p5,p6, layout_matrix = rbind(c(1,2),c(3,4),c(5,6)))
```

These boxplots show the outliers very clearly in the x and y-coordinates as well as the distance and time between relocations. In previous steps we removed some measurements with NA's for latitude and longitude to be able to convert the dataframe to a spatial dataframe. As a result, the time between relocations is not 100% accurate. We will therefore use only x-coordinates, y-coordinates and distance between relations to clean our dataset. Tresholds for the x and y-coordinates are based purely on these boxplots. To determine the treshold for the distance we take a look at the distribution of steplengths. 

```{r}
# plot histograms of distance between location points
par(mfrow=c(3, 2))
for (i in 1:length(id(movementTrajectories))){
  hist(movementTrajectories[[i]]$dist, xlim = c(0, 5000), ylim = c(0, 1500), breaks = seq(0, 1000000, by = 100), xlab = "distance (m)", main = paste("Distance between relocations by", id(movementTrajectories)[i]))
}
```
Looking at the distribution of steplengths and taking into account the consideration that roe deer generally do not travel great distances, a moved distance of 1500m or higher in one relocation is considered unrealistic. 

```{r}
# tresholds
xMin <- 650000
xMax <- 700000
yMin <- 5092500
yMax <- 5107500
distMax <- 1500

# remove outliers based on x-coordinates and distance between relocations
groupsCleaned <- groups %>% filter(x < xMax) %>% filter(x > xMin) %>% filter(y < yMax) %>% filter(y > yMin) %>% filter(dist < distMax)

# cast back to ltraj 
movementTrajectoriesCleaned <- dl(groupsCleaned)
```

For visualization purposes we create Spatial Lines of the recorded trajectories without outliers.

```{r}
# select distinct animals
animalNames <- distinct(groupsCleaned, id)[[1]]

for (animal in animalNames){
  # get location data from nth animal
  locationdata <- groupsCleaned[(groupsCleaned$id == animal),]
  
  # create spatial line
  assign(paste0("traject",animal), createSpatialLine(locationdata = locationdata, proj4dataframe = proj4zone32, proj4line = proj4WGS84, lineId = animal))
}

```

We can now visualize the cleaned movement patterns of our beloved roe deer!
```{r}
m <- leaflet() %>% 
  setView(lng=11.0367, lat=46.0331, zoom = 11) %>%
  # baselayers
  addTiles(group = 'Street map') %>%  
  addProviderTiles('Stamen.Terrain', group = 'Terrain') %>% 
  addProviderTiles('Esri.WorldImagery', group = 'Satellite') %>%
  # overlay groups
  addPolylines(data = trajectAgostino, weight = 1, opacity = 0.8, color = "red", group = 'Agostino') %>% 
  addPolylines(data = trajectAlessandra, weight = 1, opacity = 0.8, color = "blue", group = 'Alessandra') %>% 
  addPolylines(data = trajectDaniela, weight = 1, opacity = 0.8, color = "orange", group = 'Daniela') %>% 
  addPolylines(data = trajectDecimo, weight = 1, opacity = 0.8, color = "yellow", group = 'Decimo') %>% 
  addPolylines(data = trajectSandro, weight = 1, opacity = 0.8, color = "purple", group = 'Sandro') %>% 
  addLayersControl(
    baseGroups = c('Street map', 'Terrain', 'Satellite'),
    overlayGroups = c('Agostino', 'Alessandra', 'Daniela', 'Decimo', 'Sandro'),
    options = layersControlOptions(collapsed = FALSE)
  )
m 
```
Note! The data might still contain a few wrongly recorded points (we are not sure whether Agostino really visited his aunt in Stravino for a coffee). However, we are not justified to remove these potential outliers on the basis of our previously defined cut-off point.

Boxplots allow for further assessment of the cleaned data. Looks a lot better, doesn't it?
```{r}
# boxplots
p1 <- ggplot(groupsCleaned, aes(x=id, y=x, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("x-coordinates of locations")
p2 <- ggplot(groupsCleaned, aes(x=id, y=y, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("y-coordinates of locations")
p3 <- ggplot(groupsCleaned, aes(x=id, y=dist, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Distance between relocations")
p4 <- ggplot(groupsCleaned, aes(x=id, y=abs.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Absolute angle between relocations")
p5 <- ggplot(groupsCleaned, aes(x=id, y=rel.angle, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Relative angle between relocations")
p6 <- ggplot(groupsCleaned, aes(x=id, y=dt, fill = id)) + geom_boxplot(alpha=0.2) + theme(legend.position = "none") + ggtitle("Time between relocations")

grid.arrange(p1,p2,p3,p4,p5,p6, layout_matrix = rbind(c(1,2),c(3,4),c(5,6)))
```

The data is ready for the analyses!

Before we start our movement analysis, we will first have a look at the reliability of our data.
The altitude measurement of the GPS data needs to be validated through a comparison with the DEM. GPS data are not reliable if its mean altitude differs significantly from the DEM values at corresponding spatial points. 

A boxplot allows us to have a first look at the difference in height measurement between DEM and GPS.

```{r}
# calculate DEM height values at point locations for which GPS height data is available.
heightDEM <- extract(DEM, as.data.frame(groupsCleaned[,c("x","y")]), method='simple', na.rm=TRUE, df=TRUE)
heightDEM$animal <- unlist(groupsCleaned[,"id"])
colnames(heightDEM) <- c("ID", "rasterVal", "animal")

heightData <- data.frame(group = rep(c("DEM", "GPS"), each = nrow(heightDEM)), height = c(heightDEM$rasterVal, groupsCleaned$height))

# visualize
boxplot(height ~ group, heightData, main = "Height boxplot GPS and DEM", ylab = "Height")
```
Ooh! The GPS data contains large outliers (the alps are not even 20 000m high!). Did Allessandro also have a coffee in outer space?

A paired t-test is performed to assess the reliability the measurements.
```{r}
# perform a paired t-test to compare mean GPS and DEM height
tTest <- t.test(heightDEM$rasterVal, groupsCleaned$height, paired = TRUE, alternative = "two.sided")
print("Results of the paired t-test:")
tTest
```

The paired t-test confirms that there is a significant difference (p-value < 2.2e-16) in mean altitude between the GPS height measurement and the DEM height values. The mean difference between the two measurements amounts to 41 meters. This confirms our suspicion that the GPS altitude measurements are not always reliable. Therefore, we will use DEM altitude measurement for our further analysis.  

Now we extract the landuse values from the recorded trajectories. 

```{r}
# extract landuse values of recorded trajectories
landuseTracks <- extract(landUse, as.data.frame(groupsCleaned[,c("x","y")]), method='simple', na.rm=TRUE, df=TRUE)
landuseTracks$animal <- unlist(groupsCleaned[,"id"])
colnames(landuseTracks) <- c("ID", "rasterVal", "animal")
```

## Simulations
In the next section we will start with the simulations of animal trajectories. We will use the simm.crw function of the adehabitatLT package for the simulations. We wrapped this function in the simulateMovement function so that it is easy to specify the number of simulations and save all simulation in an convenient dataformat. The simulations start at the same location as the respective animal, and will have the same length.

```{r}
# number of simulations per animal
nSimulations <- 2   

# perform simulations
simulations <- simulateMovement(trajectories = movementTrajectoriesCleaned, nSimulations = nSimulations, proj4 = proj4zone32)

# plot example of simulated trajectory
plot(DEM)
lines(simulations[[1]][[1]][[1]])
```

```{r}
# extract altitude and landuse values of simulations 
simulationElevations <- extractRasterValues(DEM, simulations)
simulationLandUse <- extractRasterValues(landUse, simulations)
```

```{r}
# combine recorded and simulated height data in one matrix
simElCombi <- combineSimTrack(heightDEM, animalNames, simulationElevations)
simLandCombi <- combineSimTrack(landuseTracks, animalNames, simulationLandUse)
```

## Comparison recorded and simulated trajectories

### Elevation

```{r}
# plot the profiles of the simulated profiles and the trajectory
plotProfiles(simElCombi, animalNames)
```

```{r}
# variables for boxplot 
xlab <- "Trajectory"
ylab <- "Elevation (m)"
title <- "Comparison of recorded and simulated GPS trajectories of"

# visualize the difference between recorded and simulated data in boxplots
boxplotlist <- boxplotProfiles(simElCombi, animalNames, xlab, ylab, title, addName = TRUE) 

# arrage boxplots
grid.arrange(boxplotlist[[1]],boxplotlist[[2]],boxplotlist[[3]],boxplotlist[[4]],boxplotlist[[5]], layout_matrix = rbind(c(1,2), c(3,4),c(5,6)))
```

```{r}
# calculate summary statistics for the recorded and simulated trajectories
summaryStats <- sumStats(simElCombi, animalNames)

# create empty vectors
xAxSim <- as.integer()
meansRec <- as.numeric()
meansSim <- as.numeric()
sdRec <- as.numeric()
sdSim <- as.numeric()

# store the summary statistics in a vector for plotting
for (animal in 1:length(animalNames)){
  meansRec <- c(meansRec, summaryStats[[animal]][1,1])
  meansSim <- c(meansSim, summaryStats[[animal]][1,2:ncol(summaryStats[[animal]])])
  sdRec <- c(sdRec, summaryStats[[animal]][3,1])
  sdSim <- c(sdSim, summaryStats[[animal]][3,2:ncol(summaryStats[[animal]])])
  
  # create an index vector for the x-axis
  for (col in 1:length(2:ncol(summaryStats[[animal]]))){
    xAxSim <- c(xAxSim, animal)
  }
}

# plot
par(mfrow=c(1,2), xpd = TRUE)
plot(xAxSim, meansSim, col = c('grey'), main = 'Mean elevation', xlab = 'animal number', ylab = 'elevation (m)', pch = 19)
points(meansRec, col = 'red', pch = 19)
plot(xAxSim, sdSim, col = c('grey'), ylim = c(100,650), main = 'Standard deviation elevation', xlab = 'animal number', ylab = 'elevation (m)', pch = 19)
points(sdRec, col = 'red', pch = 19)
legend(-0.8,0, c("Recorded", "Simulated"), col = c("red", "grey"), pch = 19, bty = 'n')
par(mfrow=c(1,1), xpd = FALSE)

```

### Landuse
```{r}
# plot elevation in GGplot met simulations/recorded erop?
# r = raster(xmn=621150,xmx=694170,ymn=5090490,ymx=5109210, nrow = 624, ncol = 2434)
# DEMmasked <- mask(r, DEM)
# plot(DEMmasked)

library(rasterVis)
myTheme=rasterTheme(region=rev(brewer.pal('YlGn', n=9)))
levelplot(DEM, main = "Altitude of recorded and simulated roe deer locations", margin = FALSE, par.settings=myTheme)

           
```

```{r}
# add column to legend with shorter description
landUseLegend
LABEL2split <- strsplit(as.character(landUseLegend$LABEL2), split = " ")

# add legend to land use matrix for recorded and simulated data
simLandCombiDFLegend <- addLegend(simLandCombi, legend = landUseLegend, animalNames = animalNames)

# loop over the matrices for the different animals
for (animal in 1:length(animalNames)){
 
  # group based on ind (recorded / number of simulations)
  groupedLandUseAnimal <- as.data.frame(simLandCombiDFLegend[[animal]])
  groupedLandUseCount <- groupedLandUseAnimal %>% group_by(ind, LABEL2) %>% summarise(count = n())
  
  # create histogram
  assign(paste0("h",animal),ggplot(data=groupedLandUseCount, aes(x=LABEL2, y=count, fill=ind)) +
    geom_bar(stat="identity", position=position_dodge()) + ggtitle(paste("Land use of", animalNames[animal])) + xlab(NULL) + theme(axis.text.x = element_text(angle = 15, hjust = 1)))
}

grid.arrange(h1,h2, layout_matrix = rbind(c(1,1),c(2,2)))
grid.arrange(h3,h4, layout_matrix = rbind(c(1,1),c(2,2)))
grid.arrange(h5, layout_matrix = rbind(c(1,1),c(2,2)))

```


```{r}
# disconnect database
dbDisconnect(movementDB)
```

## References
Urbano, F., & Cagnacci, F. (2014). Spatial database for GPS wildlife tracking data. Springer International Publishing.